---
title: "smRNA report"
author: "Ming Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
    toc: yes
    toc_float:
      collapsed: no
    keep_md: false
  word_document:
    toc: yes
  pdf_document:
    toc: yes
params:
  input_dir: ""
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(knitr))

knitr::opts_chunk$set(fig.width  = 12, 
                      fig.height = 8, 
                      fig.path   = "Figures/",
                      eval       = TRUE,
                      echo       = FALSE,
                      cache      = FALSE,
                      prompt     = FALSE,
                      tidy       = FALSE,
                      comment    = NA,
                      message    = FALSE,
                      warning    = FALSE,
                      eval       = T,
                      rownames.print = FALSE)
```


```{r load pkg}
library(hiseqr)
library(knitr)
library(ggplot2)
library(ggthemes)
library(ggrepel)
library(fishualize)
library(patchwork)
library(dplyr)
library(tidyr)
library(glue)
library(ggpubr) # for pie chart
library(fishualize)
```


```{r functions}
library(limma)
library(readr)
# zscore
read_count <- function(x) {
  lapply(x, function(f) {
    readr::read_delim(
      f, "\t",
      col_names = c("name", "length", "count", "extra"),
      col_types = readr::cols()) %>%
      dplyr::mutate(
        sample = gsub(".count.csv", "", basename(f)),
        group = basename(dirname(dirname(f))),
        index = basename(dirname(f)),
        zscore = limma::zscore(count, dist="chisq", df=1),
      )
  }) %>%
    dplyr::bind_rows()
}
```


## 1. Methods

small RNA reads were first mapped to  
  - 1 small structural RNAs (rRNAs, snRNAs, snoRNAs)  
  - 2 miRNA hairpin    
  
then, remove the reads (shorter than 23nt and larger than 29nt)  

The rest reads were mapped to  
  - 3 TE consensus  
  - 4 piRNA clusters  
  - 5 genome

## 2. Results

```{r}
prj_dir  <- params$input_dir
prj_dir  <- normalizePath(prj_dir)
# prj_dir  <- normalizePath("..")
smp_name <- basename(prj_dir)
```

### 2.1 Quality control

1. Trimming adapters (>18nt) 
2. Searching for the UMI+barcodes 
3. Removing PCR_dup (based on UMI+barcodes)  

`total`:     The raw reads    
`rm_ad`:     The number of reads, after trimming adapters (>18nt)   
`too_short`: The number of reads shorter than 18nt  
`no_umi`:    The number of reads, do not contains proper UMI+barcode  
`pcr_dup`:   The number of PCR duplicates    
`clean`:     The number of clean reads, for down-stream analysis  

```{r, eval=FALSE}
hiseqr::to_DT(df)
```


```{r , fig.width=8, fig.height=5, eval=FALSE}
# pie
dfx <- df %>%
  tidyr::pivot_longer(names_to = "group", values_to = "count", 
                      c("too_short", "no_umi", "pcr_dup", "clean")) %>%
  dplyr::select(id, total, group, count) %>%
  dplyr::mutate(pct   = round(count / total * 100, 1),
                label = paste0(group, "\n", pct, "%"))

dfx %>%
  ggpubr::ggpie("count", label = "label", fill = "group", lab.pos = "out") + 
  scale_fill_fish(discrete = TRUE, option = "Scarus_tricolor")
```


### 2.2 Distribution of reads

Check the number of reads in `smRNA`, `miRNA`, `size_exclude`, `TE`, `piRNA_cluster`,
`genome` and `unmap`:

```{r load data}
f  <- "../11.stat/fx_stat.reads.csv"
group_list <- c("04.smRNA", "05.miRNA", "06.size_exclude",  "07.te", 
                "08.piRNA_cluster", "09.genome", "10.unmap")
df <- read.csv(f) %>%
  dplyr::select(sample, group, all) %>% 
  dplyr::filter(group %in% c("00.raw_data", group_list)) %>%
  dplyr::rename(count = all) %>%
  dplyr::mutate(million = round(count / 1e6, 2)) 

hiseqr::to_DT(df)
```

```{r , fig.width=8, fig.height=5}
library(ggpubr)
df %>% 
  dplyr::filter(group %in% group_list) %>%
  ggpubr::ggpie("count", label = "group", fill = "group", lab.pos = "out") + 
  scale_fill_fish(discrete = TRUE, option = "Scarus_tricolor")
```


### 2.3 Proportion of 1U and 10A in reads

Calculate the distribution of reads with: `U` at position 1 and `A` at position 10.

So the reads were split into four groups: 

`U1_A10`, U1 and A10,
`U1_B10`, U1 and not A10, 
`V1_A10`, not U1, A10 
`V1_B10`, not U1, not A10

```{r, fig.width=8, fig.height=4}
df1 <- readr::read_csv(f) %>%
  dplyr::filter(group %in% group_list) %>%
  tidyr::pivot_longer(names_to = "u1a10", values_to = "count", 
                      c("U1_A10", "U1_B10", 
                        "V1_A10", "V1_B10"))

bar_plot(df1, x = "count", y = "group", label = NA, 
         group = "u1a10", direction = "horizontal", position = "fill") +
  scale_fill_fish(option = "Scarus_tricolor", discrete = TRUE)
```


### 2.4 Length distribution of reads

```{r}
f_list <- list.files("..", "*fragsize.csv$", full.names = TRUE, recursive = TRUE)
```

#### 2.4.1 Raw reads

```{r}
f1 <- f_list[grepl("00.raw_data", f_list)]
if(length(f1) >= 1) {
  df1 <- read.csv(f1[1])
}
```

```{r, fig.width=6, fig.height=3}
hiseqr::bar_plot_lendist(df1) +
  ggtitle("00.raw_data")
```


#### 2.4.2 Map to TE consensus


```{r}
# count
c_list <- list.files("../07.te/", "*.count.txt$", 
                     full.names = TRUE, recursive = TRUE)
df_te  <- read_count(c_list)
# order unique
te_rank <- df_te %>%
  dplyr::filter(index %in% "unique") %>%
  dplyr::arrange(desc(zscore)) %>%
  dplyr::pull(name)

# re-level df_te
df_te   <- df_te %>%
  dplyr::mutate(
    name  = forcats::fct_relevel(name, te_rank),
    index = forcats::fct_relevel(index, c("unique", "multi", "both")))
p_te    <- df_te %>%
   hiseqr::bar_plot(x = "zscore", y = "name", label = NA, group = "index", direction = "horizontal") +
  facet_grid(~index)
```

```{r, fig.width=5, fig.height=15}
p_te
```



`unique`: unique mapped reads;    
`multi`: multi mapped reads;  
`both`: all mapped reads;   

```{r}
f4 <- f_list[grepl("07.te", f_list)]
f4 <- f4[grepl("unique|multi|both", f4)]
f4 <- f4[! grepl("U1_A10", f4)]
f4 <- f4[! grepl("overlap", f4)]

df4 <- lapply(f4, function(i) {
  iname <- basename(dirname(i))
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=8, fig.height=4} 
hiseqr::bar_plot_lendist(df4) +
  facet_wrap(~group, ncol = 2, scales = "free")
```

Check the distribution of `U1_A10` for `unique` mapped reads:

```{r}
f4 <- f_list[grepl("07.te", f_list)]
f4 <- f4[grepl("U1_A10", f4)]
f4 <- f4[! grepl(".unmap.", f4)]
f4 <- f4[basename(dirname(dirname(f4))) == "unique"]

df4 <- lapply(f4, function(i) {
  iname <- gsub(".fragsize.csv", "", basename(i))
  iname <- gsub(smp_name, "", iname)
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```


```{r, fig.width=8, fig.height=4} 
hiseqr::bar_plot_lendist(df4) +
  facet_wrap(~group, ncol = 2, scales = "fixed")
```


#### 2.4.3 Map to piRNA_clusters

```{r}
# count
c_list <- list.files("../08.piRNA_cluster/", "*.count.txt$", 
                     full.names = TRUE, recursive = TRUE)
df_piRC  <- read_count(c_list)
# order unique
piRC_rank <- df_piRC %>%
  dplyr::filter(index %in% "unique") %>%
  dplyr::arrange(desc(zscore)) %>%
  dplyr::pull(name)

# re-level df_te
df_piRC   <- df_piRC %>%
  dplyr::mutate(
    name  = forcats::fct_relevel(name, piRC_rank),
    index = forcats::fct_relevel(index, c("unique", "multi", "both")))
p_piRC   <- df_piRC %>%
   hiseqr::bar_plot(x = "zscore", y = "name", label = NA, group = "index", direction = "horizontal") +
  facet_grid(~index)
```

```{r, fig.width=5, fig.height=15}
p_piRC
```


`unique`: unique mapped reads;    
`multi`: multi mapped reads;  
`both`: all mapped reads;   

```{r}
f6 <- f_list[grepl("08.piRNA_cluster", f_list)]
f6 <- f6[grepl("uniq|multi|both", f6)]
f6 <- f6[! grepl("multi_k2", f6)]
f6 <- f6[! grepl(".unmap.", f6)]
f6 <- f6[! grepl("U1_A10", f6)]

df6 <- lapply(f6, function(i) {
  iname <- basename(dirname(i))
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=8, fig.height=4} 
hiseqr::bar_plot_lendist(df6) +
  facet_wrap(~group, ncol = 2, scales = "free")
```

Check the distribution of `U1_A10` for `unique` mapped reads:

```{r}
f5 <- f_list[grepl("08.piRNA_cluster", f_list)]
f5 <- f5[grepl("U1_A10", f5)]
f5 <- f5[! grepl(".unmap.|overlap", f5)]
f5 <- f5[basename(dirname(dirname(f5))) == "unique"]

df5 <- lapply(f5, function(i) {
  iname <- gsub(".fragsize.csv", "", basename(i))
  iname <- gsub(smp_name, "", iname)
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=8, fig.height=4} 
hiseqr::bar_plot_lendist(df5) +
  facet_wrap(~group, ncol = 2, scales = "fixed")
```



#### 2.4.4 smRNA and miRNA reads
  
`both`: all mapped reads;   

```{r}
f2 <- f_list[grepl("04.smRNA|05.miRNA", f_list)]
f2 <- f2[grepl("uniq|multi|both", f2)]
f2 <- f2[! grepl("U1_A10|.unmap.|overlap", f2)]

df2 <- lapply(f2, function(i) {
  iname <- basename(dirname(dirname(i)))
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=10, fig.height=3} 
hiseqr::bar_plot_lendist(df2) +
  facet_wrap(~group, ncol = 2, scales = "free")
```

#### 2.4.5 size not match piRNAs [<23, >29]

```{r}
f3 <- f_list[grepl("06.size_exclude", f_list)]
df3 <- lapply(f3, function(i) {
  iname <- basename(dirname(i))
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=6, fig.height=3} 
hiseqr::bar_plot_lendist(df3) +
  facet_wrap(~group, ncol = 2, scales = "free")
```


#### 2.4.6 Map to genome

`unique`: unique mapped reads;    
`multi`: multi mapped reads;  
`unique_multi`: all mapped reads;   

```{r}
f6 <- f_list[grepl("09.genome", f_list)]
f6 <- f6[grepl("uniq|multi|both", f6)]
f6 <- f6[! grepl("multi_k2|U1_A10|overlap", f6)]

df6 <- lapply(f6, function(i) {
  iname <- basename(dirname(i))
  read.csv(i) %>%
    dplyr::mutate(group = iname)
}) %>%
  dplyr::bind_rows()
```

```{r, fig.width=8, fig.height=4} 
hiseqr::bar_plot_lendist(df6) +
  facet_wrap(~group, ncol = 2, scales = "free")
```





